Recursive Solutions to First-Order Model Counting

### 2022-02-15 notes/notes.tex
Describe search

\paragraph{Things I might need to explain.}
\begin{itemize}
\item atom, (positive/negative) literal, constant, predicate, variable, literal variable, clause, unit clause
\item $\Vars$, $\Vars(c) = \Vars(L) \cup \Vars(C)$
\item $\Doms$ on both formulas and clauses. $\Doms(c) = \Imm \delta_c$, and $\Doms(\phi) = \bigcup_{c \in \phi} \Doms(c)$.
\item maybe: notation for partial function, notation for powerset, notation for disjoint union, domain size
\item $\WMC$, $\wwp$, $\wwn$, $\Imm$
\item two parts: compilation and inference.
\item introduce and use arrows for bijections, injections, set inclusions, etc.
\item $\emptyset$ for an empty partial map.
\item notation for variable substitution for literal and constraint sets: $S[x/V]$, where $S$ is the set, $V$ is a set of variable, and $x$ is the new constant or variable. Note that for constraint sets this could mean that a variable and a constant need to switch places.
\item During inference, there is a domain size map $\size\colon \mathscr{D} \to \mathbb{N}_0$.
\end{itemize}

@@

\paragraph{TODO.}
\begin{itemize}
\item maybe add a (vertex) labelling function and its codomain (and use it in the algorithm)? But it's only used once...
\item explain the connection to the algebraic formalism (or just don't use it). % We will also use algebraic notation to describe (parts of) circuits: circuittype_parameters(nodes pointed to by the outedges)
\item define `an FCG for a formula'
\item need an example
\item vertex or node, $v$ or $n$?
\item what notation for lists (a.k.a. sequences) am I going to use? $\oplus$ for concatenation (of both two lists and an element to a list), $\in$ for (in-order) enumeration, $[]$ for an empty list, $[x]$ for a list with one element, $L$ for the length of the list.
\end{itemize}

@@

\begin{algorithm}[t]
  \caption{The search algorithm}
  \label{alg:search}
  \SetKwData{solutions}{solutions}
  \SetKwProg{Fn}{Function}{:}{}
  \SetKwFunction{applyGreedyRules}{applyGreedyRules}
  \SetKwFunction{applyNonGreedyRules}{applyNonGreedyRules}
  \SetKwFunction{put}{put}
  \SetKwFunction{get}{get}
  \SetKwFunction{emptyy}{empty}
  \SetKwFunction{updateCache}{updateCache}
  \KwData{a set of greedy rules $\Gamma$}
  \KwData{a set of non-greedy rules $\Delta$}
  \KwData{a formula $\phi_0$}
  \KwResult{all found FCGs for $\phi_0$ are in the set \solutions}
  $\solutions \gets \emptyset$\;
  $C_0 \gets \emptyset$\;
  $(G_0, C_0, L_0) \gets \applyGreedyRules{$\phi_0$, $C_0$}$\;
  \lIf{$L_0 = []$}{$\solutions \gets \solutions \cup \{\, G_0 \,\}$}
  \Else{
    $q \gets \text{an empty queue of states}$\;
    $q.\put{$(G_0, C_0, L_0)$}$\;
    \While{{\bf not} $q.\emptyy{}$}{
      \ForEach{state $(G, C, L) \in \applyNonGreedyRules{$q.\get{}$}$}{
        \lIf{$L = []$}{$\solutions \gets \solutions \cup \{\, G \,\}$}
        \lElse{$q.\put{$(G, C, L)$}$}
      }
    }
  }
  \Fn{\applyGreedyRules{$\phi$, $C$}}{
    \ForEach{rule $r \in \Gamma$}{
      $S \gets r(\phi)$\;
      \If{$S \ne \emptyset$}{
        $(G, L) \gets \text{any element of } S$\; % TODO: explain why this is valid for greedy rules
        \lIf{$G = {\normalfont \texttt{null}}$}{\Return{\applyGreedyRules{the only element of $L$, $C$}}}
        \Else{
          $(V, s, N^+) \gets G$\;
          $C \gets \updateCache(C, \phi, s)$\;
          % TODO: continue here
        }
      }
    }
    \Return{$({\normalfont \texttt{null}}, C, [\phi])$}\;
  }
\end{algorithm}

>>>>

\paragraph{Things I might need to explain.}
\begin{itemize}
\item atom, (positive/negative) literal, constant, predicate, variable, literal variable, clause, unit clause
\item $\Vars$, $\Vars(c) = \Vars(L) \cup \Vars(C)$
\item $\Doms$ on both formulas and clauses. $\Doms(c) = \Imm \delta_c$, and $\Doms(\phi) = \bigcup_{c \in \phi} \Doms(c)$.
\item maybe: notation for partial function, notation for powerset, notation for disjoint union, domain size
\item $\WMC$, $\wwp$, $\wwn$, $\Imm$
\item two parts: compilation and inference.
\item introduce and use arrows for bijections, injections, set inclusions, etc.
\item $\emptyset$ for an empty partial map.
\item notation for variable substitution for literal and constraint sets: $S[x/V]$, where $S$ is the set, $V$ is a set of variable, and $x$ is the new constant or variable. Note that for constraint sets this could mean that a variable and a constant need to switch places.
\item During inference, there is a domain size map $\size\colon \mathscr{D} \to \mathbb{N}_0$.
\item $\sqcup$ for both sets and functions
\end{itemize}

@@

\paragraph{TODO.}
\begin{itemize}
\item maybe add a (vertex) labelling function and its codomain (and use it in the algorithm)? But it's only used once...
\item explain the connection to the algebraic formalism (or just don't use it). % We will also use algebraic notation to describe (parts of) circuits: circuittype_parameters(nodes pointed to by the outedges)
\item define `an FCG for a formula'
\item vertex or node, $v$ or $n$?
\item what notation for lists (a.k.a. sequences) am I going to use? $\oplus$ for concatenation (of both two lists and an element to a list), $\in$ for (in-order) enumeration, $[]$ for an empty list, $[x]$ for a list with one element, $|L|$ for the length of the list, and $h : t$ to denote a list with first element (a.k.a. head) $h$ and remaining list (a.k.a. tail) $t$.
\item use Greek letters for compilation rules
\end{itemize}

@@

\begin{algorithm}[t]
  \caption{The (main part of the) search algorithm}
  \label{alg:search}
  \SetKwProg{Fn}{Function}{:}{}
  \SetKwData{solutions}{solutions}
  \SetKwFunction{applyGreedyRules}{applyGreedyRules}
  \SetKwFunction{applyNonGreedyRules}{applyNonGreedyRules}
  \SetKwFunction{put}{put}
  \SetKwFunction{get}{get}
  \SetKwFunction{emptyy}{empty}
  \KwIn{a formula $\phi_0$}
  \KwResult{all found FCGs for $\phi_0$ are in the set \solutions}
  $\solutions \gets \emptyset$\;
  $C_0 \gets \emptyset$\;
  $(G_0, C_0, L_0) \gets \applyGreedyRules{$\phi_0$, $C_0$}$\;
  \lIf{$L_0 = []$}{$\solutions \gets \{\, G_0 \,\}$}
  \Else{
    $q \gets \text{an empty queue of states}$\;
    $q.\put{$(G_0, C_0, L_0)$}$\;
    \While{{\bf not} $q.\emptyy{}$}{
      \ForEach{state $(G, C, L) \in \applyNonGreedyRules{$q.\get{}$}$}{
        \lIf{$L = []$}{$\solutions \gets \solutions \cup \{\, G \,\}$}
        \lElse{$q.\put{$(G, C, L)$}$}
      }
    }
  }
\end{algorithm}

\begin{algorithm}
  \caption{Functions used in \cref{alg:search} for applying compilation rules}
  \label{alg:apply}
  \SetKwProg{Fn}{Function}{:}{}
  \SetKwData{newStates}{newStates}
  \SetKwFunction{applyGreedyRules}{applyGreedyRules}
  \SetKwFunction{applyNonGreedyRules}{applyNonGreedyRules}
  \SetKwFunction{applyGreedyRulesToAllFormulas}{applyGreedyRulesToAllFormulas}
  \SetKwFunction{mergeFcgs}{mergeFcgs}
  \SetKwFunction{updateCache}{updateCache}
  \KwData{a set of greedy rules $\Gamma$}
  \KwData{a set of non-greedy rules $\Delta$}
  \Fn{\applyGreedyRules{$\phi$, $C$}}{
    \ForEach{rule $r \in \Gamma$}{
      $S \gets r(\phi)$\;
      \If{$S \ne \emptyset$}{
        $(G, L) \gets \text{any element of } S$\;
        \lIf{$G = {\normalfont \texttt{null}}$}{\Return{\applyGreedyRules{the only element of $L$, $C$}}}
        \Else{
          $(V, s, N^+) \gets G$\;
          $C \gets \updateCache(C, \phi, s)$\;
          \Return{\applyGreedyRulesToAllFormulas{$G$, $C$, $L$}}\;
        }
      }
    }
    \Return{$({\normalfont \texttt{null}}, C, [\phi])$}\;
  }
  \Fn{\applyGreedyRulesToAllFormulas{$G$, $C$, $L$}}{
    $(V, s, N^+) \gets G$\;
    $N^+(s) \gets []$\;
    $L' \gets []$\;
    \ForEach{formula $\phi \in L$}{
      $(G', C, L'') \gets \applyGreedyRules{$\phi$, $C$}$\;
      $L' \gets L' \oplus L''$\;
      \lIf{$G' = {\normalfont \texttt{null}}$}{$N^+(s) \gets N^+(s) \oplus [\star]$}
      \Else{
        $(V', s', N') \gets G'$\;
        $V \gets V \sqcup V'$\;
        $N^+ \gets N^+ \sqcup N'$\;
        $N^+(s) \gets N^+(s) \oplus [s']$\;
      }
    }
    \Return{$((V, s, N^+), C, L')$}\;
  }
  \Fn{\applyNonGreedyRules{$s$}}{
    $(G, C, L) \gets s$\;
    $\phi : T \gets L$\;
    $(G', C', L') \gets \text{a copy of } s$\;
    $\newStates \gets []$\;
    \ForEach{rule $r \in \Delta$}{
      \ForEach{$(G'', L'') \in r(\phi)$}{
        \lIf{$G'' = {\normalfont \texttt{null}}$}{$\newStates \gets \newStates \oplus \applyNonGreedyRules{$(G', C', L'')$}$}
        \Else{
          $(V, s, N^+) \gets G''$\;
          $C' \gets \updateCache{$C'$, $\phi$, $s$}$\;
          $(G'', C', L'') \gets \applyGreedyRulesToAllFormulas{$G''$, $C'$, $L''$}$\;
          \lIf{$G' = {\normalfont \texttt{null}}$}{$\newStates \gets \newStates \oplus [(G'', C', L'' \oplus T)]$}
          \lElse{$\newStates \gets \newStates \oplus [(\mergeFcgs{$G'$, $G''$}, C', L'' \oplus T)]$}
        }
      }
      $(G', C', L') \gets \text{a copy of } s$\;
    }
    \Return{\newStates}\;
  }
\end{algorithm}

\begin{algorithm}
  \caption{Helper functions used by \cref{alg:apply}}
  \label{alg:helpers}
  \SetKwProg{Fn}{Function}{:}{}
  \SetKwFunction{mergeFcgs}{mergeFcgs}
  \SetKwFunction{updateCache}{updateCache}
  \Fn{\updateCache{$C$, $\phi$, $v$}}{
    \lIf{$\#\phi \not\in \dom(C)$}{\Return{$C \cup \{\, \#\phi \mapsto (\phi, v) \,\}$}}
    \lIf{there is no $(\phi', v') \in C(\#\phi)$ such that $v' = v$}{$C(\#\phi) \gets (\phi, v) \oplus C(\#\phi)$}
    \Return{$C$}\;
  }
  \Fn{\mergeFcgs{$G = (V, s, N^+)$, $G' = (V', s', N')$, $r = s$}}{
    \lIf{$r$ has label \textsc{Ref}}{\Return{\normalfont \texttt{null}}}
    \ForEach{$t \in N^+(r)$}{
      \If{$t = \star$}{
        replace $t$ with $s'$ in $N^+(r)$\;
        \Return{$(V \sqcup V', s, N^+ \sqcup N')$}\;
      }
      $G'' \gets \mergeFcgs{$G$, $G'$, $t$}$\;
      \lIf{$G'' \ne {\normalfont \texttt{null}}$}{\Return{$G''$}}
    }
    \Return{\normalfont \texttt{null}}\;
  }
\end{algorithm}

\paragraph{TODO}
\begin{itemize}
\item maybe `applyAllRules' is a better name
\item explain the `tail' part of the algorithm
\end{itemize}

Note: At the end, \texttt{mergeFcgs} will never return \texttt{null} because there is going to be at least one $\star$ in $G$ and the function will find it.
